The Cost of Undisturbed Landscapes

### 2020-05-03 projects/asparagus/doc/paper_clean.tex
updates asparagus paper_clean.tex

\documentclass[final, 3p, times]{elsarticle} %review, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,

@@

    \section{Introduction} \label{sec:introduction}
    Future, low carbon electricity systems should largely rely on renewable energy technologies (RET), such as solar
    photovoltaics (PV) or wind turbines, according to a substantial body of literature~\citep{Rodriguez2015}.\todo{references}
    For central Europe, it is typically found that system-cost\footnote{System cost are the sum of fuel and emission
    cost,operation and maintenance cost, and investment cost} minimizing systems to rely on large shares of onshore
    wind power~\citep{Rodriguez2015}\todo{references}.
    Such undertakings frequently implement upper bounds on the deployment of RET, which are thought to reflect the
    available potential for RET erection~\citep{references}\todo{references}.
    These potentials, in turn, are usually derived from surveys of land available and suitable for the erection of RET\@.
    The typical methodology is to exclude areas that are considered not suited because of insufficient resource
    availability or because of technical or legal restrictions such as minimum distances to settlements.
    Areas not excluded are then considered as feasible for RET deployment \citep[e.g.][]{McKenna2015, Bosch2017, Ryberg2020}.
    \todo[inline]{refer to excluded areas as black, feasible areas as white}
    However, researchers are increasingly aware that such a 'black-or-white'-approach fails to account for more nuanced
    issues such as the social acceptance of interference with landscapes.
    Indeed, several contributions sought to improve on such shortcomings, for example by participatory
    modelling~\citep{Hoeltinger2016}, by xxx~\citep{Jaeger2016} or xxx~\citep{Harper2019}\todo[inline]{add mini review paragraph}

    In contrast, the work of~\cite{Drechsler2011} % Drechsler et al. 2011: Combining spatial modelling and choice experiments ...
    \todo[inline]{mention later paper by Drechsler} \cite{Drechsler2017}
    \todo[inline]{refer to negative externalities not included in system cost}

    We improve on this work by xxx. \todo[inline]{explain what we do also in contrast to Drechsler et al.}

    \subsection{Framework}
    Similar to~\cite{Drechsler2011}, we regard the problem as one of welfare optimization, in which we seek to minimize
    the social cost (including, but not limited to system cost) of low carbon energy systems.
    Our framework is best understood by considering a benevolent social planner who is to decide on the deployment of
    RET\@.
    Such a social planner faces a trade-off between
    (i) (potential) negative externalities of a RET at the local level (recall that externalities from emissions are
    already included in system cost), such as the visual impact on landscapes and
    (ii) (potential) positive impact of a RET through lowering system cost (compared to its best alternative).

@@

    For illustration, consider figure~\ref{figure:social-optimum}, which depicts the effect of deviating from a
    system-cost minimizing RET allocation by substituting some share of technology $A$ with its best alternative,
    technology $B$.\footnote{Here, we consider the interesting case in which in system-cost minimum a RET causes higher
    external cost than it saves in system cost.}
    We let $\alpha$ denote the share of technology $A$ substituted by the best available alternative technology $B$.
    At $\alpha=0$, we are in the system cost minimum, while the 'harm' $h$ from otherwise unaccounted external cost is
    high.
    Now suppose that we can lower external cost by substituting $A$ with $B$, i.e.\ harm $h$ is decreasing in $\alpha$.
    A benevolent social planner will substitute $A$ with $B$ as long as the resulting increase in system cost is lower
    than the corresponding decline in harm.
    \bigskip
    \todo[inline]{the following subsection must be overhauled}
    Within this framework, we quantify the opportunity cost of wind power ($s$ in Figure~\ref{figure:social-optimum})
    for the case of Austria in the work at hand.
    We chose Austria as our case study, as the government has comitted itself to meeting \SI{100}{\percent} of
    electricity consumption from renewable sources (on annual balance, excluding industry own-consumption and system
    services) by 2030.
    This policy makes Austria an ideal case for studying the substitution of RET while keeping total renewable
    electricity generation constant.
    Moreover, the bulk of additional renewable electricity must come either from wind power or solar photovoltaics, as
    the other renewables potentials are largely exploited already (see section~\ref{section:}), which simplifies our
    analysis further.
    In the context of Austria, as in many other European countries, least-cost renewable electricity systems are largely
    based on wind power~\citep{Rodriguez2015}. \todo[inline]{references}
    Harm is inflicted by local external effects of wind turbines, such as flickering, noise, negative impacts on
    wildlife, or (perceived) disturbance of landscapes.
    \cite{Mattmann2016} identifies the latter as a main trigger of opposition against wind power at the local level.
    For a review of wind power externalities, see~\cite{Zerrahn2017}.
    \todo[inline]{1) motivate Austrian case, 2) make clear that paper only looks at opportunity cost of wind power --
    cost of undisturbed landscapes, NOT at harm, which is assessed in various ways by many papers, though not for
    Austria}

@@

    \section{Data and Methods} \label{sec:data-and-methods}
    To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system
    model \emph{medea}, summarized in section~\ref{subsec:medea}, which we instantiate with scenario assumptions
    (see section~\ref{subsec:data}) and observed data of the year 2016 (see section~\ref{subsec:scenarios}).

    For determining the opportunity cost of wind power, we proceed as follows:
    \begin{enumerate}
        \item We derive the unrestricted system cost-minimizing deployment of wind and solar power,
        \item we restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind power)
        and observe net system cost $c_{net}$ for Austria (calculated as total system cost including air pollution cost
        net of trade balance).\label{enumerate:restrict}
        \item We repeat step~\ref{enumerate:restrict} till no wind power can be deployed.
        \item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by
        the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.\
        \[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
        where $\Delta$ is the difference operator.
    \end{enumerate}

% ***** medea *****

    \subsection{Power system model \emph{medea}} \label{subsec:medea}
    We make use of the power system model \emph{medea} to simulate (dis)investment in and hourly operation of the
    prospective Austrian (and German) power system.

    The model is cast as a linear optimization seeking to minimize total system cost, which consist of fuel and emission
    cost, quasi-fixed and variable operation and maintenance (O\&M) cost, the costs of investment in energy generation,
    storage, and transmission assets, and (potential) cost of non-served load.

    The system is required to meet exogenous and inelastic demand for electricity and heat in any hour of the modeled
    year.
    Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and
    transmission units.
    Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's
    electrical efficiency, the electricity loss per unit of heat production, and the back pressure coefficient.
    Electricity generation from intermittent sources (wind, water, solar) is subject to exogenous hourly generation
    profiles, which are scaled according to total installed capacities.
    Electricity from these sources can be curtailed at no additional cost (free disposal).

    \todo{Energy storage}
    Electricity can be stored in reservoir and pumped hydro storages or batteries.
    Generation from hydro storage plants is constrained by turbine capacity and energy contained in reservoirs of
    limited size.
    Hydro reservoirs are filled by inflows or by pumping (pumped storages only).
    To better capture operational differences, we model daily, weekly and seasonal reservoir and pumped storage plants
    separately.

    \todo{Electricity exchange}
    Electricity can be exchanged between Austria and Germany.

    \todo{Ancillary services}
    To ensure a stable and secure operation of the electricity system, power plants must provide ancillary services
    (e.g. frequency control, voltage support).
    We assume that this requires generators with an installed capacity of at least 21 GW2 to be operational (either
    generating or pumping in case of pumped hydro storages) at any point in time.

    For a detailed, mathematical description please refer to \ref{sec:medea}.

    Data processing is implemented in python, the optimization model is based on GAMS. Running the model on an Intel
    i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver, takes 10--15 minutes, depending on the model
    parameters.
    \todo[inline]{perfect foresight, perfect competition, reflects energy-only market}

@@

    \subsubsection{Scenario Assumptions for Austria}\label{subsubsec:assumptions-austria}
    Austria has set itself the goal of generating \SI{100}{\percent} of electricity consumption from renewable sources
    on annual balance by 2030.
    However, industry own consumption and system services, which currently account for \SI{10}{\percent} of annual
    electricity consumption, are exempt.
    Hence, in our analysis we set the policy goal to generating at least \SI{90}{\percent} of annual electricity
    consumption from renewable sources.\footnote{Please note that this policy goal does not imply that the generated
    electricity must actually be consumed.
    Hence, we consider curtailed electricity as contributing to the policy goal.}
    The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity
    annually from renewable sources.
    Consequently, we have set prospective electricity consumption accordingly.

    New hydro power plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity
    generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
    We add corresponding generation capacities to reach these targets.

    The remainder is projected to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind turbines
    (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.

    The latter two RET are the only low-carbon electricity generation technologies with significant expansion potential.
    Austrians ruled out the use of nuclear power in a 1978 referendum~\citep{Pelinka1983}.
    Biomass lacks ecological and economical sustainability~\citep{Erb2018}, while available hydro power potentials were
    already largely exploited~\citep{Walder2014} even before adding another \SI{5}{\tera\watt\hour} of annual generation.

    As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for
    endogenous investment in wind and solar power, without enforcing announced targets.

    \subsubsection{Scenario Assumptions for Germany}\label{subsubsec:assumptions-germany}
    Germany has announced specific capacity targets for several power generation technologies.
    Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to
    recommendations by~\cite{WSB2019}, and a further expansion of renewable electricity generation in line
    with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
    We do, however, expect a lifting of the \SI{52}{\giga\watt} limit on solar PV\@.

    \begin{table}[H]
        \caption{Initial Generation Capacities}
        \centering
        \begin{tabulary}{0.55\textwidth}{lcc}
            \toprule
            Technology & Austria & Germany \\
            & GW & GW \\
            \midrule
            Wind Onshore & $2.6$ & $90.8$ \\
            Wind Offshore & $0.0$ & $15.0$ \\
            Solar PV & $1.1$ & $73.0$ \\
            Run-of-river Hydro & $6.7$ & $4.5$ \\
            Biomass & $0.725$ & $8.4$ \\
            Lignite & $0.0$ & $11.4$ \\
            Coal & $0.0$ & $14.0$ \\
            Natural Gas & $3.9$ & $24.2$ \\
            Mineral Oil & $0.2$ & $3.5$ \\
            Heat Pump & $1.0$ & $1.0$ \\
            Gas Boiler & $3.9$ & $25.5$ \\ \bottomrule
        \end{tabulary}
        \label{tab:instcaps}
    \end{table}

    \subsection{Energy supply}
    We represent 27\todo{compare with table!} dispatchable energy conversion and storage technologies that are expected
    to be operational by 2030.
    In addition to exogenously set, initally installed capacities (see~\ref{tab:instcaps}), the model can endogenously
    add further generation capacities that are compatible with stated policy objectives.
    All technologies can also be decommissioned, so that we adopt a long-run perspective on the power system.

    Assumed technology parameters are summarized in Table~\ref{itm:techcost}.

    \begin{table*}[h]
        \centering
        \begin{threeparttable}
            \caption{Technology Input Parameters}\label{itm:techcost}
            %\begin{longtable}{p{2.9cm} p{1.5cm} p{1.5cm} p{1.6cm} p{1.5cm} p{1.0cm} p{1.2cm} p{1.0cm}}
            \begin{tabulary}{0.5\textwidth}{lccccccl}
                \toprule
                & \multicolumn{2}{c}{Capital Cost} & \multicolumn{2}{c}{O\&M-Cost} & & & \\ \cmidrule(lr){2-3} \cmidrule(lr){4-5}
                Technology & Power & Energy & Quasi- & Variable & Life- & Efficiency & Source    \\
                & & & fixed & & time & &           \\
                & \EUR/kW & \EUR/kWh & \EUR/MWa & \EUR/MWh & a & & \\
                \midrule
                %\multicolumn{8}{c}{\emph{Renewable Energy Technologies}}\\ \midrule
                Wind Onshore & $1040$    & NA & $12600$   & $1.35$     & $30$ & NaN & 1 \\
                Wind Offshore & $1930$    & NA & $36053$   & $2.70$     & $30$ & NaN & 1 \\
                %Solar PV       & $630$ & NaN & $8844$ & $0.00$ & $40$ & NaN & 9 \\
                PV Rooftop & $870$     & NA & $10815$   & $0.00$     & $40$ & NaN & 1 \\
                PV Open-space & $330$     & NA & $6380$    & $0.00$     & $40$ & NaN & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Thermal Energy Conversion}}\\ \midrule
                Lignite Adv & -- & NA & $40500$   & $0.85$     & $40$ & 0.439 & \cite{OekoInstitut2017} \\
                Coal SC & -- & NA & $25000$   & $6.00$     & $40$ & 0.425 & 1 \\
                Coal USC & -- & NA & $31000$   & $2.90$     & $40$ & 0.485 & 1 \\
                Nat Gas ST & $400$     & NA & $15000$   & $3.00$     & $30$ & 0.407 & 1 \\
                Nat Gas GT & $590$     & NA & $19500$   & $4.40$     & $25$ & 0.420 & 1 \\
                Nat Gas CC & $880$     & NA & $29300$   & $4.40$     & $25$ & 0.610 & 1 \\
                Oil ST & $400$     & NA & $6000$    & $3.00$     & $30$ & 0.396 & 1 \\
                Oil GT & $378$     & NA & $8068$    & $4.50$     & $25$ & 0.410 & 1 \\
                Oil CC & $800$     & NA & $25000$   & $4.00$     & $25$ & 0.470 & 1 \\
                Biomass & $3400$    & NA & $140000$  & $4.50$     & $25$ & 0.298 & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Energy Storage Technologies}}\\ \midrule
                Hydro Reservoir & -- & -- & $X$       & $X$       & $80$ & $0.9$  & X \\
                Hydro PSP & -- & -- & $X$       & $X$       & $80$ & $0.9^2$ & X\\
                Battery Li-Ion & $1930$    & $321$ & $36053$   & $2.70$    & $30$ & NaN & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Heat Generation Technologies}}\\ \midrule
                Resistive Heater & $X$      & $X$   & $X$       & $X$       & $25$  & NaN & X \\
                Heatpump & $500$     & NaN & $36053$   & $2.70$    & $30$ & NaN & 1 \\
                Natural gas boiler & $60$   & NA & $1950$    & $1.00$    & $20$ & 0.90 & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Electricity Transmission}}\\ \midrule
                Transmission & $1930$    & NaN & $36053$   & $2.70$    & $30$ & NaN & 1 \\
                \bottomrule
            \end{tabulary}

@@

    Hourly generation from non-dispatchable technologies solar photovoltaics, wind turbines and run-of-river hydro is
    assumed to follow hourly profiles as observed in 2016.
    \todo[inline]{more details on derivation of load profiles for Austria, source for Germany, scaling}

    Similarly, hourly inflows into hydro reservoirs (including pumped storage reservoirs) are inferred from weekly
    reservoir levels in combination with hourly generation and pumping time series published by~\cite{ENTSO-E2020}.
    However, all storage technologies (including hydro reservoir and pumped storage plant) are charged and dispatched
    endogenously.
    \todo[inline]{improve description of methodology for inflow approximation}

>>>>

\documentclass[review, 3p, times]{elsarticle} %final, twocolumn, 5p, 12pt, authoryear, twocolumn, 5p,

@@

    \section{Introduction} \label{sec:introduction}
    Future, low carbon energy systems are likely to largely rely on electricity from renewable energy technologies
    (RET), such as solar photovoltaics (PV) or wind turbines, according to a substantial body of
    literature~\citep[e.g.][]{Becker2014,Delucchi2011,Fernandes2014,Gils2017,Jacobson2011,Schmidt2016}.
    \todo{references not good for statement}
    A broad stream of papers~\citep[e.g.][]{Brown2018,Rodriguez2015,Schlachtberger2018,Scholz2017} investigates
    conditions under which total system cost, conceived as the sum of fuel and emission cost, operation and maintenance
    cost, and investment cost, can be minimized.
    In this context, available resource potentials are an important boundary condition for the determination of such
    least-cost system configurations.
    Fortunately, such resource potentials are assessed by numerous scholars at various levels.
    Frequently, these assessments combine indicators of the wind resource with technical or legal restrictions on land
    available for RET deployment~\citep[e.g.][]{Bosch2017,McKenna2015,Ryberg2020}.
    However, there is an increasing awareness that such approaches are not a proper reflection of deployable RET
    capacites, as a more nuanced analysis would have to account for the social acceptance of RET.
    Opposition to wind turbines, for example, is often based on perceived negative interference with
    landscapes~\cite{Mattmann2016}, potential harm to wildlife~\cite{Loss2013, Voigt2015, Wang2015}, or
    noise~\cite{Wang2015a}.
    Zerrahn~\cite{Zerrahn2017} provides an excellent survey of the literature on wind power externalities.

    Based on this insight, several contributions sought to take social acceptance into account when assessing RET
    potentials.
    Höltinger et al.\ \cite{Hoeltinger2016}, for example, relied on participatory modeling to identify uncontested wind
    power potentials in Austria.
    Jäger et al.\ \cite{Jaeger2016} \todo[inline]{brief description of paper}
    Harper et al.\ \cite{Harper2019} develop a spatially resolved multi-criteria analysis including technological, legal,
    and social constraints to determine sites suitable for wind turbine erection.
    % mention minimum distances to settlements

    In contrast to the determination of such `augmented potentials' trying to identify sites that are certainly suited
    for RET deployment, Drechsler et al.\ \cite{Drechsler2011} argue that a welfare-optimal spatial allocation of RET
    needs to consider social costs and benefits.
    Social conflict would then be mitigated by selecting sites with the lowest welfare loss.
    To demonstrate their approach, Drechsler et al.\ conduct a case study for a region in Germany, where they survey
    preferences of inhabitants to explicitly consider the trade-off between \emph{social cost} arising from negative
    externalities of wind power and electricity generation cost of wind power due to spatially varying wind conditions.
    Some of the authors studied implications of welfare optimal allocation of wind turbines also in more recent
    contributions.
    Salomon et al. (2020)~\citep{Salomon2020} study welfare-optimal rules on minimum distances between wind turbines and
    human settlements.
    Drechsler et al. (2017)~\citep{Drechsler2017} study welfare-optimal allocation of wind turbines and solar PV.\todo[inline]{expand on this}

    We expand on this work by conceptualising welfare optimal RET deployment in more general settings, so that we can
    account for the opportunity cost of RET substitution (arising from different system integration cost of RET), as
    well as spatial differences in endowments with renewable resources and negative externalities from RET depending on
    spatially diverse characteristics, as we lay out in section~\ref{subsec:framework} below.
    \todo[inline]{how? accounting for everything could be done by spatially resolved model that has info on resource
    availability and on negative impact for each location.
    Alternatively, use model for opportunity cost only and calc deviation against mean resource?}

    To undergird our proposed methodology, we use a numerical power system model (see
    section~\ref{sec:data-and-methods}) to investigate the trade-off between deploying different renewable energy
    technologies in terms of system cost, which constitute an important, though incomplete part of social cost.
    We do this to underline the importance of RET substitution and the corresponding opportunity costs in the
    identification of spatial allocations of RET as well as in power system planning in general.
    In the latter context, we also contribute to the analysis of near-optimal power system
    configurations~\citep[see e.g.][]{Neumann2019,Schlachtberger2017}.

    % What do we want to show?
    % We embed our demonstration in a wider framework suited to expand renewable energy technologies at minimal social cost,
    % thereby mitigating potential social conflict over the transition towards low-carbon energy systems.
    % The current contribution demonstrates that:
    % - there is a tradeoff between different RET

    \subsection{Framework} \label{subsec:framework}
    Similar to~\cite{Drechsler2011}, we regard the problem as one of welfare optimization, in which we seek to minimize
    the social cost (including, but not limited to system cost) of low carbon energy systems.
    Our framework is best understood by considering a benevolent social planner who is to decide on the deployment of
    RET\@.
    The social planner faces a trade-off between
    (i) (potential) negative externalities of a RET at the local level (recall that externalities from emissions are
    already included in system cost), such as the visual impact on landscapes and
    (ii) (potential) positive impact of a RET through lowering system cost (compared to its best alternative).

@@

    For illustration, consider figure~\ref{figure:social-optimum}, which depicts the effect of deviating from a
    system-cost minimizing RET allocation by substituting some share $\alpha$ of technology $A$ with its best alternative,
    technology $B$.\footnote{Here, we consider the interesting case in which in system-cost minimum a RET causes higher
    external cost than it saves in system cost.}
    We let $\alpha$ denote the share of technology $A$ substituted by the best available alternative technology $B$.
    At $\alpha=0$, we are in the system cost minimum, while the 'harm' $h$ from otherwise unaccounted external cost is
    high.
    Now suppose that we can lower external cost by substituting $A$ with $B$, i.e.\ harm $h$ is decreasing in $\alpha$.
    A benevolent social planner will substitute $A$ with $B$ as long as the resulting increase in system cost is lower
    than the corresponding decline in harm.

    \subsection{Case study}
    Based on this framework, we use the case of Austria to quantify the opportunity cost of wind power (line $s$ in
    figure~\ref{figure:social-optimum}).
    We do not set out to quantify harm (i.e.\ negative external effects) induced by wind power.
    While such an endeavor is beyond the scope of this paper, it is a frequent undertaking in the
    literature~\citep[e.g.][]{Droees2016, Gibbons2015, Heintzelmann2017, Jensen2018, Kussel2019, Lang2014, Sims2008, Sunak2016, Vyn2014},
    even though an assessment of negative wind turbine externalities has to the best of our knowledge not been conducted
    for Austria yet.

    We chose Austria as our case to study, as the government has committed itself to meeting \SI{100}{\percent} of
    electricity consumption from renewable sources (on annual balance, excluding industry own-consumption and system
    services) by 2030.
    This policy makes Austria an ideal case for studying the substitution of RET while keeping total renewable
    electricity generation constant.
    Moreover, in Austria the bulk of additional renewable electricity must come either from wind power or solar
    photovoltaics, as other renewables potentials are largely exploited already (see
    section~\ref{subsubsec:assumptions-austria:}), which simplifies our analysis further.

    In the context of Austria, we find a least-cost renewable electricity system to be largely based on wind power,
    which allows us to quantify the opportunity cost of wind power in terms of its best alternative solar PV.
    Corresponding results are given in section~\ref{sec:results-discussion}.

@@

    \section{Data and Methods} \label{sec:data-and-methods}
    To investigate the cost of undisturbed landscapes (i.e. the opportunity cost of wind power) we use the power system
    model \emph{medea}, summarized in section~\ref{subsec:medea}, which we instantiate with scenario assumptions
    and observed data of the year 2016 (see section~\ref{subsec:data}).
    For determining the opportunity cost of wind power, we proceed as follows:
    \begin{enumerate}
        \item We derive the unrestricted system cost-minimizing deployment of wind and solar power,
        \item we restrict deployment of wind power by a small margin (so that the next best RET substitutes for wind power)
        and observe net system cost $c_{net}$ for Austria (calculated as total system cost including air pollution cost
        net of trade balance).\label{enumerate:restrict}
        \item We repeat step~\ref{enumerate:restrict} till no wind power can be deployed.
        \item Finally, we approximate the opportunity cost of wind power (at given wind power capacity $w$) $OC_w$ by
        the change in $c_{net}$ in response to a change in wind power capacity $w$ deployed, i.e.\
        \[OC_w = \frac{\Delta c_{net}}{\Delta w}\]
        where $\Delta$ is the difference operator.
    \end{enumerate}

% ***** medea *****
    \subsection{Power system model \emph{medea}} \label{subsec:medea}
    We make use of the power system model \emph{medea} to simulate (dis)investment in and hourly operation of the
    prospective Austrian (and German) power system.
    The model is cast as a linear optimization seeking to minimize total system cost, which consist of fuel and emission
    cost, quasi-fixed and variable operation and maintenance (O\&M) cost, the costs of investment in energy generation,
    storage, and transmission assets, and (potential) cost of non-served load.
    In terms of economics, the model reflects a perfectly competitive energy-only market with fully price-inelastic
    demand and perfect foresight of all actors.

    The system is required to meet exogenous and inelastic demand for electricity and heat in any hour of the modeled
    year.
    Energy supply, in turn, is constrained by available installed capacities of energy conversion, storage, and
    transmission units.
    Co-generation units convert fuel to heat and power subject to a feasible operating region defined by the unit's
    electrical efficiency, the electricity loss per unit of heat production, and the back pressure coefficient.
    Electricity generation from intermittent sources (wind, run-of-river hydro, solar) is subject to exogenous hourly
    generation profiles, which are scaled according to total installed capacities.
    Electricity from these sources can be curtailed at no additional cost (free disposal).

    Electricity can be stored in reservoir and pumped hydro storages or batteries.
    The capacity of hydro storages can not be expanded, as we assume existing potentials to be exhausted.
    Battery capacity, on the other hand, can be added endogenously.
    Generation from storages is constrained by installed capacity and energy contained in storage.
    Hydro reservoirs are filled by exogenously given inflows.
    Pumped hydro storages and batteries can actively store electricity for later use.
    To better capture operational differences of hydro storage units, we model daily, weekly and seasonal reservoir and
    pumped storage plants separately.
    To ensure a stable and secure operation of the electricity system, ancillary services (e.g. frequency control,
    voltage support) are required.
    We model ancillary service needs as a minimum requirement on spinning reserves operating at any point in time.
    Thus, we assume that ancillary services can be provided by thermal power plant, run-of-river hydro plant or any
    type of storage technology.
    We do not model transmission and distribution grids within Austria or Germany.
    We do, however, account for the congestion management scheme introduced by German and Austrian authorities in 2018,
    that limits cross-border electricity trade.
    \todo[inline]{is this elegant? Should i mention that investment in x-border capacity is allowed but does not occur?}
    For a detailed, mathematical description please refer to \ref{sec:medea}.

    Data processing is implemented in python, while the optimization model is based on GAMS.
    Running the model on an Intel i7-8700 machine with 16 GB RAM, using CPLEX 12.9 as a solver, takes 10--15 minutes,
    depending on the model parameters.

@@

    \subsubsection{Scenario Assumptions for Austria}\label{subsubsec:assumptions-austria}
    The Austrian government has set itself the goal of generating \SI{100}{\percent} of electricity consumption from
    renewable sources on annual balance by 2030~\citep{Regierungsprogramm2020}.
    However, industry own consumption and system services, which currently account for about \SI{10}{\percent} of annual
    electricity consumption, are exempt.
    The government plans to achieve this goal by generating an additional \SI{27}{\tera\watt\hour} of electricity
    annually from renewable sources\footnote{Please note that this policy goal does not imply that the generated
    electricity must actually be consumed. Hence, we count curtailed electricity as contributing to the policy goal.}.
    New hydro power plants are thought to contribute \SI{5}{\tera\watt\hour} annually, while additional electricity
    generation from biomass is envisaged to account for \SI{1}{\tera\watt\hour} annually.
    We add corresponding generation capacities to reach these targets.
    The government projects the remainder to come from solar PV (\SI{11}{\tera\watt\hour} annually) and onshore wind
    turbines (\SI{10}{\tera\watt\hour} annually)~\cite{Regierungsprogramm2020}.
    As we are interested in determining the opportunity cost of wind power versus its best alternative, we allow for
    endogenous investment in wind and solar power, without enforcing announced targets.
    Other low-carbon energy technologies are not feasible at large scale in the Austrian context.
    The use of nuclear power was banned in a 1978 referendum~\citep{Pelinka1983}.
    Biomass lacks ecological as well as economical sustainability~\citep{Erb2018}, and hydro power potentials were
    already largely exploited~\citep{Walder2014} even before government plans of adding another \SI{5}{\tera\watt\hour}
    in annual hydro power generation.
    Initial generation capacities for Austria are given in table~\ref{tab:instcaps}.

    \subsubsection{Scenario Assumptions for Germany}\label{subsubsec:assumptions-germany}
    Germany has announced specific capacity targets for several power generation technologies.
    Following these announcements, we anticipate an end to nuclear power generation, a (partial) coal exit according to
    recommendations by the `coal commission'~\cite{WSB2019}, and a further expansion of renewable electricity generation
    in line with the German Renewable Energy Sources Act~\cite{noauthor_renewable_2017}.
    We do, however, expect the \SI{52}{\giga\watt} solar PV limit to be lifted.
    The corresponding capacity assumptions are displayed in table~\ref{tab:instcaps}.

    \begin{table}[H]
        \caption{Initial Generation Capacities}
        \centering
        \begin{tabulary}{0.55\textwidth}{lcc}
            \toprule
            Technology & Austria & Germany \\
            & GW & GW \\
            \midrule
            Wind Onshore & $2.6$ & $90.8$ \\
            Wind Offshore & $0.0$ & $15.0$ \\
            Solar PV & $1.1$ & $73.0$ \\
            Run-of-river Hydro & $6.9$ & $4.5$ \\
            Biomass & $0.725$ & $8.4$ \\
            Lignite & $0.0$ & $11.4$ \\
            Coal & $0.0$ & $14.0$ \\
            Natural Gas & $3.9$ & $24.2$ \\
            Mineral Oil & $0.2$ & $3.5$ \\
            Heat Pump & $1.0$ & $1.0$ \\
            Gas Boiler & $3.9$ & $25.5$ \\ \bottomrule
        \end{tabulary}
        \label{tab:instcaps}
    \end{table}

    \subsection{Energy supply}
    We represent 27\todo{compare with table!} dispatchable energy conversion and storage technologies that are expected
    to be operated in 2030.
    In addition to given, initally installed capacities (see~\ref{tab:instcaps}), the model can endogenously add further
    generation capacities that are compatible with stated policy objectives.
    All technologies can also be decommissioned, so that we adopt a long-run perspective on the power system.
    Parameters of admissible technologies are summarized in Table~\ref{itm:techcost}.

    \begin{table*}[h!t]
        \centering
        \begin{threeparttable}
            \caption{Technology Input Parameters}\label{itm:techcost}
            %\begin{longtable}{p{2.9cm} p{1.5cm} p{1.5cm} p{1.6cm} p{1.5cm} p{1.0cm} p{1.2cm} p{1.0cm}}
            \begin{tabulary}{0.5\textwidth}{lccccccl}
                \toprule
                & \multicolumn{2}{c}{Capital Cost} & \multicolumn{2}{c}{O\&M-Cost} & & & \\ \cmidrule(lr){2-3} \cmidrule(lr){4-5}
                Technology & Power & Energy & Quasi- & Variable & Life- & Efficiency & Source    \\
                & & & fixed & & time & &           \\
                & \EUR/kW & \EUR/kWh & \EUR/MWa & \EUR/MWh & a & & \\
                \midrule
                %\multicolumn{8}{c}{\emph{Renewable Energy Technologies}}\\ \midrule
                Wind Onshore & $1040$    & NA & $12600$   & $1.35$     & $30$ & NaN & 1 \\
                Wind Offshore & $1930$    & NA & $36053$   & $2.70$     & $30$ & NaN & 1 \\
                %Solar PV       & $630$ & NaN & $8844$ & $0.00$ & $40$ & NaN & 9 \\
                PV Rooftop & $870$     & NA & $10815$   & $0.00$     & $40$ & NaN & 1 \\
                PV Open-space & $330$     & NA & $6380$    & $0.00$     & $40$ & NaN & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Thermal Energy Conversion}}\\ \midrule
                Lignite Adv & -- & NA & $40500$   & $0.85$     & $40$ & 0.439 & \cite{OekoInstitut2017} \\
                Coal SC & -- & NA & $25000$   & $6.00$     & $40$ & 0.425 & 1 \\
                Coal USC & -- & NA & $31000$   & $2.90$     & $40$ & 0.485 & 1 \\
                Nat Gas ST & $400$     & NA & $15000$   & $3.00$     & $30$ & 0.407 & 1 \\
                Nat Gas GT & $590$     & NA & $19500$   & $4.40$     & $25$ & 0.420 & 1 \\
                Nat Gas CC & $880$     & NA & $29300$   & $4.40$     & $25$ & 0.610 & 1 \\
                Oil ST & $400$     & NA & $6000$    & $3.00$     & $30$ & 0.396 & 1 \\
                Oil GT & $378$     & NA & $8068$    & $4.50$     & $25$ & 0.410 & 1 \\
                Oil CC & $800$     & NA & $25000$   & $4.00$     & $25$ & 0.470 & 1 \\
                Biomass & $3400$    & NA & $140000$  & $4.50$     & $25$ & 0.298 & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Energy Storage Technologies}}\\ \midrule
                Hydro Reservoir & -- & -- & $X$       & $X$       & $80$ & $0.9$  & X \\
                Hydro PSP & -- & -- & $X$       & $X$       & $80$ & $0.9^2$ & X\\
                Battery Li-Ion & $1930$    & $321$ & $36053$   & $2.70$    & $30$ & NaN & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Heat Generation Technologies}}\\ \midrule
                Resistive Heater & $X$      & $X$   & $X$       & $X$       & $25$  & NaN & X \\
                Heatpump & $500$     & NaN & $36053$   & $2.70$    & $30$ & NaN & 1 \\
                Natural gas boiler & $60$   & NA & $1950$    & $1.00$    & $20$ & 0.90 & 1 \\ \midrule
                %\multicolumn{8}{c}{\emph{Electricity Transmission}}\\ \midrule
                Transmission & $1930$    & NaN & $36053$   & $2.70$    & $30$ & NaN & 1 \\
                \bottomrule
            \end{tabulary}

@@

    Generation from non-dispatchable technologies solar photovoltaics, wind turbines and run-of-river hydro is assumed
    to follow hourly generation profiles as observed in 2016.
    German solar and wind power profiles are sourced from Open Power System Data~\citep{opsd2019}.
    For Austria, solar and wind power generation profiles were derived by dividing solar and wind power generation
    timeseries reported by Open Power System Data~\cite{opsd2019} by installed capacities published by \cite{Biermayr}.
    Similarly, generation profiles for run-of-river hydro power were derived based on generation data reported by
    \cite{ENTSOE2020}.

    Data on inflows into Austrian hydro reservoirs is not publicly available.
    However, ENTSO-E's transparency platform~\cite{ENTSOE2020} reports data on hydro storage filling levels, electricity
    generation from hydro reservoirs, and pumping by pumped hydro storages.
    Based on this data, we have approximated inflows as the part of change in (weekly) hydro storage fill levels not
    explained by (upsampled hourly) pumping or generation.
    To arrive at hourly inflows, we have interpolated our weekly estimates by piecewise cubic hermite interpolation.
    Although inflows to hydro reservoirs are given exogenous, all storage technologies (including hydro reservoir and
    pumped storage plant) are charged and dispatched endogenously.
